---
title: "Backup"
output: html_notebook
---

# 8. Expected Number of Clonal Mutations

## 8.1 Derived by Bozic et al. Using a Birth-Death Model

We have the formula for expected number of clonal passenger mutations from `Bozic et al., 2016` as follow:

$$
\overline{m_c} = \frac{\frac{1}{R_0} * \mu}{1-\frac{1}{R_0}}
$$

```{r}
meanClonalBozic <- function(R0, mu) {
  delta <- 1 / R0
  mc <- (delta * mu) / (1 - delta)
  return(mc)
}

meanClonalBozic.df <- function(R0_values, mu_values) {
  # e.g. mu_values <- seq(0,0.1,by = 0.0001)
  mean <- 1:length(mu_values) %>%
    mclapply(function(x) {
      1:length(R0_values) %>%
        lapply(function(y) {
          meanClonalBozic(R0_values[y], mu_values[x])
        })
    }, mc.cores = detectCores())

  mcdf <- expand.grid(R0 = paste0("", R0_values), mu = paste0("", mu_values))
  vs_slope <- c()

  mcdf$mc <- unlist(mean)
  return(mcdf)
}
```

We can plot the mean clonal mutations calculated by the formula derived by Bozic for a set of $R_0$ and $\mu$ values in a 3D axes. 

```{r}
df_mcBozic <- meanClonalBozic.df(seq(1.01, 7.01, by = 0.01), seq(0.01, 0.11, by = 0.01))

plot_ly(
  x = ~ df_mcBozic$R0, y = ~ df_mcBozic$mu,
  z = ~ df_mcBozic$mc, type = "mesh3d"
) %>%
  layout(scene = list(
    xaxis = list(nticks = 4, title = "R0", range = c(1, 7)),
    yaxis = list(nticks = 4, title = "mu", range = c(0, 0.1)),
    zaxis = list(
      type = "log", ticks = 4,
      title = "mean # of clonal variants"
    )
  ))
```


## 8.2 Derived by Our Model

We can use the following function to calculate the expected clonal mutation number with given parameter.

$$
E(Clonal)=\sum_{i=0}^{n}iP(Clonal=i)
$$

```{r}
meanClonal <- function(n, R0, mu, maxMuGen, maxFS, maxClonal) {
  df <- pmf.P_clonal_new(n, R0, mu, maxMuGen, maxFS, maxClonal)
  sum <- 0
  for (i in 1:(maxClonal + 1)) { # omit 0 to simplify the calculation process
    sum <- sum + i * df[i + 2, 2]
  }
  return(sum)
}

meanClonal.df <- function(n, R0_values, mu_values, maxMuGen, maxFS, maxClonal) {
  # e.g. mu_values <- seq(0,0.1,by = 0.0001)
  mean <- 1:length(mu_values) %>%
    mclapply(function(x) {
      1:length(R0_values) %>%
        lapply(function(y) {
          meanClonal(n, R0_values[y], mu_values[x], maxMuGen, maxFS, maxClonal)
        })
    }, mc.cores = detectCores())

  mcdf <- expand.grid(R0 = paste0("", R0_values), mu = paste0("", mu_values))
  vs_slope <- c()

  mcdf$mc <- unlist(mean)
  return(mcdf)
}
```

We can plot the mean clonal mutations calculated by our method for a set of $R_0$ and $\mu$ values in a 3D axes. 

```{r, eval = FALSE}
df_mc <- meanClonal.df(1, seq(1.1, 1.5, by = 0.01), seq(0.01, 0.1, by = 0.01), 50, 50, 5)

plot_ly(
  x = ~ df_mc$R0, y = ~ df_mc$mu,
  z = ~ df_mc$mc, type = "mesh3d"
) %>%
  layout(scene = list(
    xaxis = list(nticks = 4, title = "R0"),
    yaxis = list(nticks = 4, title = "mu"),
    zaxis = list(type = "log", ticks = 4, title = "mean # of clonal variants")
  ))
```


# 10. Branching Process Model

## 10.1 Create Offspring of One Haplotype

```{r}
CreateOffspringOfOneHaplotype <- function(n_thisHaplotype, hap_num, ngens, R0, mu_probs) {
  p <- 1 / (R0 + 1)
  Znplus1_thisHaplotype <- rgeom(n_thisHaplotype, p)
  tot_Znplus1 <- sum(Znplus1_thisHaplotype)

  tempZnMatrix <- c()
  tempparentVector <- c()
  tempmutationVector <- c()

  if (tot_Znplus1 == 0) {
    offspringSameHaplotype <- 0
  } else {
    haplotypeOffspring <- rmultinom(1, tot_Znplus1, mu_probs)
    offspringSameHaplotype <- haplotypeOffspring[1]

    for (i in 2:length(mu_probs)) {
      if (haplotypeOffspring[i] > 0) {
        n_mutations <- i - 1
        insertMatrix <- cbind(
          matrix(0,
            nrow = haplotypeOffspring[i],
            ncol = length(1:ngens)
          ),
          matrix(1, nrow = haplotypeOffspring[i], ncol = 1)
        )
        tempZnMatrix <- rbind(tempZnMatrix, insertMatrix)
        tempparentVector <- rbind(
          tempparentVector,
          matrix(hap_num,
            nrow = haplotypeOffspring[i],
            ncol = 1
          )
        )
        tempmutationVector <- rbind(
          tempmutationVector,
          matrix(n_mutations,
            nrow = haplotypeOffspring[i],
            ncol = 1
          )
        )
      }
    }
  }

  list <- list(offspringSameHaplotype, tempZnMatrix, tempparentVector, tempmutationVector)
  names(list) <- c("offspringSameHaplotype", "tempZnMatrix", "tempparentVector", "tempmutationVector")
  return(list)
}
```

## 10.2 Get Mutation List Haplotypes

```{r}
GetMutationListHaplotypes <- function(nhaplotypes_final, parentVector, mutationVector) {
  mutationList <- c()
  maxMut <- 0
  for (i in 2:nhaplotypes_final) {
    nMuts <- mutationVector[i]
    extraMuts <- maxMut + (1:nMuts)
    mutationList[i] <- c(mutationList[parentVector[i]], extraMuts)
    maxMut <- max(extraMuts)
  }
  list <- list(mutationList, maxMut)
  names(list) <- c("mutationList", "maxMut")
  return(list)
}
```

## 10.3 Calculate Clonal Mutations

```{r}
CalculateClonalMutations <- function(finalPopSizes, nhaplotypes_final, mutationList, maxMut) {
  totFinalPopSize <- sum(finalPopSizes)
  mutFinalPopSizes <- matrix(0, nrow = maxMut, ncol = 1)
  for (i in 1:nhaplotypes_final) {
    this_list <- mutationList[i]
    k <- this_list
    mutFinalPopSizes[k, 1] <- mutFinalPopSizes[k, 1] + finalPopSizes[i]
  }
  locs_clonal <- which(mutFinalPopSizes == totFinalPopSize)
  nclonal <- length(locs_clonal)
  return(nclonal)
}
```

## 10.4 Simulate One Realization of Branching Process

```{r}
SimulateOneRealization_BranchingProcess <- function(n, R0, mu, Nmax) {
  p <- 1 / (R0 + 1)
  Zn <- data.frame(n)
  parentVector <- c(0)
  mutationVector <- c(0)
  bool_extinct <- 0

  max_mutations <- max(5, qpois(0.99, mu))
  mu_probs <- dpois(0:max_mutations, mu)
  mu_probs[length(mu_probs)] <- mu_probs[length(mu_probs)] + (1 - sum(mu_probs))

  while (TRUE) {
    nTotHaplotypes <- dim(Zn)[1]
    ngens <- dim(Zn)[2]
    Zn[1:nTotHaplotypes, ngens + 1] <- 0

    tempZnMatrix <- c()
    tempparentVector <- c()
    tempmutationVector <- c()

    for (i in 1:nTotHaplotypes) {
      list <- CreateOffspringOfOneHaplotype(Zn[i, ngens], i, ngens, R0, mu_probs)
      names(list) <- c(
        "offspringSameHaplotype", "tempZnMatrix_thisHaplo",
        "tempparentVector_thisHaplo", "tempmutationVector_thisHaplo"
      )
      Zn[i, ngens + 1] <- as.numeric(list["offspringSameHaplotype"])
      tempZnMatrix <- rbind(tempZnMatrix, list[["tempZnMatrix_thisHaplo"]])
      tempparentVector <- rbind(tempparentVector, list[["tempparentVector_thisHaplo"]])
      tempmutationVector <- rbind(
        tempmutationVector,
        list[["tempmutationVector_thisHaplo"]]
      )
    }

    Zn <- data.frame(rbind(t(matrix(Zn)), tempZnMatrix))
    parentVector <- rbind(parentVector, tempparentVector)
    mutationVector <- rbind(mutationVector, tempmutationVector)

    tot_Zn <- sum(as.numeric(Zn[, ngens + 1]))

    if (tot_Zn > Nmax) {
      break
    }
    if (tot_Zn == 0) {
      bool_extinct <- 1
      break
    }
  }

  nhaplotypes <- dim(Zn)[1]
  ngens <- dim(Zn)[2]

  list <- list(nhaplotypes, ngens, Zn, parentVector, mutationVector, bool_extinct)
  names(list) <- c("nhaplotypes", "ngens", "Zn", "parentVector", "mutationVector", "bool_extinct")
  return(list)
}
```

## 10.5 Simulate Branching Process Model

```{r}
branchingProcess <- function(n_stoch_real, bool_onlySuccessful, n, R0, mu, Nmax) {
  results_nclonal <- c()
  for (i in 0:n_stoch_real) {
    simOneList <- SimulateOneRealization_BranchingProcess(n, R0, mu, Nmax)
    names(simOneList) <- c(
      "nhaplotypes_final", "ngens_final", "Zn",
      "parentVector", "mutationVector", "bool_extinct"
    )
    if (bool_extinct == 1) {
      if (bool_onlySuccessful == 0) {
        i <- i + 1
        results_nclonal[i] <- c(0)
      }
    } else {
      i <- i + 1
      mutList <- GetMutationListHaplotypes(
        simOneList[["nhaplotypes_final"]],
        simOneList[["parentVector"]],
        simOneList[["mutationVector"]]
      )
      nclonal <- CalculateClonalMutations(
        Zn[, simOneList[["ngens_final"]]],
        simOneList[["nhaplotypes_final"]],
        mutList[["mutationList"]],
        mutList[["maxMut"]]
      )
      results_nclonal[i] <- nclonal
    }
  }
  return(results_nclonal)
}
```
